Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-01-16 manuscript.tex
added figure example, edits to tex

\documentclass[12pt]{article}
\usepackage{hyperref}
\usepackage[pdftex]{graphicx}  
\usepackage{caption}
\usepackage{natbib}
\usepackage{color}
\usepackage{authblk}
\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} }

@@

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, was identified as a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (Hopscotch) upstream of tb1 enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the Hopscotch element, led us to investigate the role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the Hopscotch element is segregating at a higher than expected frequency in a number of populations of both subspecies \emph{parviglumis} and \emph{mexicana}. Analysis of linkage disequilibrium (LD) between the Hopscotch element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, yet we find no correlation between the frequency of the Hopscotch insertion and environmental variables. Additionally, we find no evidence that the Hopscotch plays a similar role in branching architecture in teosinte as it does in domesticated maize.

\section{INTRODUCTION}

@@

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis for these changes has been a focus of maize research for a number of years. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \cite{Doebley et al 1990, Doebley and Stec 1991, Lukens and Doebley 1999}. Further studies have shown that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer et al 2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \cite{Studer et al 2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \cite{Studer et al 2011}. Furthermore, Studer and Doebley (2012) \cite{Studer and Doebley 2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. In this study we aim to characterize the distribution of the Hopscotch insertion in \emph{mexicana}, \emph{parviglumis}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the Hopscotch element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the Hopscotch element and tillering phenotypes in a population of \emph{parviglumis}.

\section{METHODS}

\subsection{Sampling and Genotyping}

We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 Zea mays ssp. mexicana (hereafter mexicana) populations, and 86 \emph{parviglumis} populations) and assessed the presence or absence of the Hopscotch insertion (Table S1, Table S2). DNA was extracted from leaf tissue using a modified CTAB approach \cite{Doyle and Doyle 1990, Maloof et al 1984}. We designed primers using PRIMER3 \cite{Rozen and Skaletsky 2000} implemented in Geneious \cite{Kearse et al 2012} to amplify the entire Hopscotch element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence (~5 kb amplification product) and absence (~300 bp amplification product) of the Hopscotch insertion. Two PCRs were performed for each individual, one with primers flanking the Hopscotch (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, 5'-TCGTTGATGCTTTGATGGATGG-3'; 
Hop R, 5'-AACAGTATGATTTCATGGGACCG-3'; and HopIntR, 5'-CCTCCACCTCTCATGAGATCC-3' (Figure S1). Homozygotes show a single band for either the Hopscotch element (~5 kb) or the absence of the element (~300 bp), while heterozygotes are three-banded, showing a band for both the presence and absence of the Hopscotch element as well as a band for the internal primer set (Figure S2). When only one PCR resolved well, we scored one allele for the individual, which explains the odd number of alleles included in our analyses. We used Phusion High Fidelity Enzyme (Finnzymes, Inc.) and the following conditions for amplifications: 98°C for 3 min, 30 cycles of 98°C for 15 s, 65°C for 30 s, and 72°C for 3 min 30 s, with a final extension of 72°C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the Hopscotch based on band size.

\subsection{Sequencing}

@@

\subsection{Genotyping Analysis}

@@

\subsection{Sequence Analysis}

@@

\subsection{Phenotyping of \emph{Zea mays.} ssp. \emph{parviglumis}}

@@

\section{Results}

\subsection{Genotyping}

@@

\subsection{Sequencing}

@@

\subsection{Evidence of introgression}

@@

\subsection{Phenotyping}

\subsubsection{Phenotyping 1}

@@

\subsection{Phenotyping 2}

@@

\section{DISCUSSION}

Adaptation occurs either due to selection on standing variation or on de novo mutations. Adaptation as a result of selection on standing variation has been well-described in a number of systems, for example, selection on lactase tolerance in humans \cite{Plantinga et al 2012, Tishkoff et al 2007}; variation at the Eda locus in three-spined stickleback \cite{Kitano et al 2008, Colosimo et al 2005}; and pupal diapause in the Apple Maggot fly \cite{Feder et al 2003}. Although the role of standing variation with respect to adaptation has been described in many systems, its importance to domestication is not as well studied.

In maize, alleles at important domestication loci (RAMOSA1, \cite{Sigmon and Vollbrecht 2010}; barren stalk1, \cite{Gallavotti et al 2004}; and grassy tillers1, \cite{Whipple et al 2011} have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \cite{Clark et al 2006, Studer et al 2011}. Studer et al. (2011) genotyped 90 accessions of teosinte (inbred and non-inbred), and provided the first evidence that the Hopscotch insertion is segregating in teosinte \cite{Studer et al 2011}. Our findings suggest that the Hopscotch insertion is more widespread in populations of teosinte than previously thought and that the presence of the insertion is not due to recent introgression. However, we find no effect of the Hopscotch insertion on branching architecture in teosinte.

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it is still segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by Studer et al. (2011) in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \cite{Studer et al 2011}. Crop to wild introgression has previously been observed at domestication loci (\cite{Ellstrand et al 1999, Zhang et al 2009, Thurber et al 2010, Baack et al 2008, Hubner et al 2012, Wilkes et al 1977, Van Heerwaarden et al 2011, Barrett 1983}), but our results are more consistent with Hufford et al. (2013) who found resistance to introgression from maize into teosinte \cite{Hufford et al 2013}. Furthermore, Hufford et al. (2013) showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize).

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the tb1 locus had introgressed into populations of teosinte. We do not have sympatric maize samples from many of our populations, however, so we cannot rule out the possibility that Hopscotch alleles in these teosinte populations are segregating in sympatric maize. On the other hand, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1. The similarity in LD in the \emph{tb1} region in populations with and without the Hopscotch provides further support for an explanation other than introgression. Additionally, assignment to maize in this region in our STRUCTURE analysis is in fact lower than the average across chromosome 1 (Figure 3, Table 4).

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the Hopscotch in these populations. In the ancient introgression scenario, recombination would have broken up linked variants surrounding the Hopscotch and we would find LD patterns similar to the ones we report here. However, it seems unlikely that there would have been ancient introgression between these populations of teosinte and their sympatric maize populations and not present-day introgression. Studies have demonstrated the presence of ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \cite{Hufford et al 2013, Ellstrand et al 2007, van Heerwaarden et al 2011}.

The lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D in the San Lorenzo population are consistent with patterns expected under neutrality. Though, deviations from expected genotype frequencies under Hardy-Weinberg equilibrium at the Hopscotch insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus. Similarly, if the Hopscotch is segregating as neutral variation in teosinte, the probability that it would have simultaneously drifted to high frequency in multiple populations is quite low; however, we find no evidence of selection or introgression at or near the \emph{tb1} locus in our sample set.

The phenotypic effects of the Hopscotch insertion in domesticated maize have been well documented \cite{Clark et al 2006, Studer et al 2011}, and Weber et al. (2007) have described its effects in partial inbred lines of teosinte \cite{Weber et al 2007}. Our study is the first to explicitly examine the phenotypic effects of the Hopscotch insertion in a population of teosinte. However, we found no significant effect of the Hopscotch on tillering index or tiller number in our phenotyping experiments, and the effect of the Hopscotch insertion in teosinte is discordant with that of maize. The lack of correlation between Hopscotch genotype and tillering index or tiller number is surprising given its effects in maize, and preliminary data collected on a grow-out of the TILs (teosinte inbred lines).

It is certainly possible that even though data from the TILs suggest an effect of the Hopscotch on tillering in teosinte, that it in fact has no effect. Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \cite{Clark et al 2006}, and perhaps we would have detected an effect of the Hopscotch on these additional phenotypes. A recent study by Studer and Doebley (2012) examined the possibility of an allelic series at the \emph{tb1} locus in teosinte by introgressing 9 separate teosinte segments spanning the \emph{tb1} locus into an isogenic maize background and investigating their effects on phenotype. However, they found evidence for an allelic series contributing to all \emph{tb1} associated phenotypes but tillering. They did find variation in tillering both between introgressed lines and between isogenic maize lines, suggesting that there may be additional loci that contribute to tillering in both maize and teosinte \cite{Studer and Doebley 2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering.

Another explanation for the lack of correlation between Hopscotch genotype and tillering could be growing conditions in out greenhouse affected growth habit of our plants. The \emph{tb1} gene is part of the shade avoidance pathway in maize, and this involvement may have presented problems for our phenotyping design \citep{Kebrom and Brutnell 2007}. Broadly, shade avoidance syndrome (SAS), refers to the morphological and physiological response of plants experiencing an increase in far-red (FR) light, which occurs during shading \cite{Kebrom and Brutnell 2007}. Under increased FR light conditions plants experience both morphological and physiological changes including reduction in branching, increased height, and increased ethylene and auxin production \cite{Ballare 1999, Smith and Whitelam 1997}. In maize, increase in shading has been shown to be correlated with changes in \emph{tb1} expression and subsequent reduction in tillering, suggesting that \emph{tb1} is involved in the shade avoidance response in maize \cite{Doebley et al 1997, Lukens and Doebley 1999}. If plants in our phenotyping experiments were grown too close together, such that they shaded each other, it could have reduced \emph{tb1} expression and tillering, which would have confounded our results.

In summary, our findings demonstrate that the Hopscotch allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the Hopscotch allele is present in these populations due to recent introgression; however, it seems unlikely that the it would have drifted to high frequency in multiple populations, and there may be another explanation for the high frequency we observe in some of our populations. The Hopscotch insertion does not appear to have any affect on tillering phenotypes in teosinte previously associated with \emph{tb1} and it is possible that variation at other loci may play a larger role in branching architecture in teosinte than in domesticated maize.

\clearpage
\bibliography{refs}
\bibliographystyle{apalike}

>>>>

\documentclass[12pt]{article}
\usepackage{hyperref}
\usepackage[pdftex]{graphicx}  
\usepackage{caption}
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{color}
\usepackage{authblk}
\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} }

@@

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (Hopscotch) upstream of tb1 enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the Hopscotch element, led us to investigate the role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the Hopscotch element is segregating at a higher than expected frequency in a number of populations of both subspecies \emph{parviglumis} and \emph{mexicana}. Analysis of linkage disequilibrium between the Hopscotch element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection.  Finally, two greenhouse experiments fail to find an important role for \emph{tb1} in controlling tillering in natural populations of \emph{parvilgumis}. \jri{Needs a concluding sentence. "Thus, we think..." or something.}

\section*{Introduction}
\jri{plase redo sections not in all caps}

@@

% SFS figure
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{SFS.png}
    \caption{This is an example figure.} 
\label{sfs_non_syn}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

\jri{This is how you cite the figure in the text. "I want to show you all Figure \ref{sfs_non_syn}." Latex takes care of the numbering.  Please add all the figures to the text, and add them in in the middle like this close to where they are referenced. Note that it puts the figure where it thinks best, which is not necessarily near this text.}

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis for these changes has been a focus of maize research for a number of years. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \cite{Doebley et al 1990, Doebley and Stec 1991, Lukens and Doebley 1999}. Further studies have shown that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer et al 2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \cite{Studer et al 2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \cite{Studer et al 2011}. Furthermore, \cite{Studer and Doebley 2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. In this study we aim to characterize the distribution of the Hopscotch insertion in \emph{mexicana}, \emph{parviglumis}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the Hopscotch element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the Hopscotch element and tillering phenotypes in a population of \emph{parviglumis}.

\section*{METHODS}

\subsection*{Sampling and Genotyping}

We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 Zea mays ssp. mexicana (hereafter mexicana) populations, and 86 \emph{parviglumis} populations) and assessed the presence or absence of the Hopscotch insertion (Table S1, Table S2). \jri{please add tables and reference same as figs; see example in Sofiane's "deleterious 282" repo on lab github page} DNA was extracted from leaf tissue using a modified CTAB approach \cite{Doyle and Doyle 1990, Maloof et al 1984}. We designed primers using PRIMER3 \cite{Rozen and Skaletsky 2000} implemented in Geneious \cite{Kearse et al 2012} to amplify the entire Hopscotch element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence (~5 kb amplification product) and absence (~300 bp amplification product) of the Hopscotch insertion. Two PCRs were performed for each individual, one with primers flanking the Hopscotch (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, 5'-TCGTTGATGCTTTGATGGATGG-3'; 
Hop R, 5'-AACAGTATGATTTCATGGGACCG-3'; and HopIntR, 5'-CCTCCACCTCTCATGAGATCC-3' (Figure S1). \jri{supplemental material should be added to this file too.} Homozygotes show a single band for either the Hopscotch element (~5 kb) or the absence of the element (~300 bp), while heterozygotes are three-banded, showing a band for both the presence and absence of the Hopscotch element as well as a band for the internal primer set (Figure S2). When only one PCR resolved well, we scored one allele for the individual, which explains the odd number of alleles included in our analyses. We used Phusion High Fidelity Enzyme (Finnzymes, Inc.) and the following conditions for amplifications: 98°C for 3 min, 30 cycles of 98°C for 15 s, 65°C for 30 s, and 72°C for 3 min 30 s, with a final extension of 72°C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the Hopscotch based on band size.

\subsection*{Sequencing}

@@

\subsection*{Genotyping Analysis}

@@

\subsection*{Sequence Analysis}

@@

\subsection*{Phenotyping of \emph{Zea mays.} ssp. \emph{parviglumis}}

@@

\section*{Results}

\subsection*{Genotyping}

@@

\subsection*{Sequencing}

@@

\subsection*{Evidence of introgression}

@@

\subsection*{Phenotyping}

\subsubsection*{Phenotyping 1}

@@

\subsection*{Phenotyping 2}

@@

\section*{DISCUSSION}

Adaptation occurs either due to selection on standing variation or on de novo mutations. Adaptation as a result of selection on standing variation has been well-described in a number of systems, for example, selection on lactase tolerance in humans \cite{Plantinga et al 2012, Tishkoff et al 2007}; variation at the \jri{italics?} Eda locus in three-spined stickleback \cite{Kitano et al 2008, Colosimo et al 2005}; and pupal diapause in the Apple Maggot fly \cite{Feder et al 2003}. Although the role of standing variation with respect to adaptation has been described in many systems, its importance to domestication is not as well studied.

In maize, alleles at important domestication loci (RAMOSA1, \cite{Sigmon and Vollbrecht 2010}; barren stalk1, \cite{Gallavotti et al 2004}; and \jri{do gene names need to be italicized?} grassy tillers1, \cite{Whipple et al 2011} have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \cite{Clark et al 2006, Studer et al 2011}. Studer et al. (2011) genotyped 90 accessions of teosinte (inbred and non-inbred), and provided the first evidence that the Hopscotch insertion is segregating in teosinte \cite{Studer et al 2011}.

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it is still segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by Studer et al. (2011) in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \cite{Studer et al 2011}. Crop to wild introgression has previously been observed at domestication loci (\cite{Ellstrand et al 1999, Zhang et al 2009, Thurber et al 2010, Baack et al 2008, Hubner et al 2012, Wilkes et al 1977, Van Heerwaarden et al 2011, Barrett 1983}), but our results are more consistent with Hufford et al. (2013) who found resistance to introgression from maize into teosinte \cite{Hufford et al 2013}. Furthermore, \jri{all citations, even citations like these need to reference the bibtex file. check out the difference betweetn cite, citet, and citep as commands.} Hufford et al. (2013) showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize).

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte. We do not have sympatric maize samples from many of our populations, however, so we cannot rule out the possibility that Hopscotch alleles in these teosinte populations are segregating in sympatric maize. On the other hand, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1. The similarity in LD in the \emph{tb1} region in populations with and without the Hopscotch provides further support for an explanation other than introgression. Additionally, assignment to maize in this region in our STRUCTURE analysis is in fact lower than the average across chromosome 1 (Figure 3, Table 4). \jri{figure references need to be as in the example using the ref command}

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the Hopscotch in these populations. In the ancient introgression scenario, recombination would have broken up linked variants surrounding the Hopscotch and we would find LD patterns similar to the ones we report here. However, it seems unlikely that there would have been ancient introgression between these populations of teosinte and their sympatric maize populations and not present-day introgression. Studies have demonstrated the presence of ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \cite{Hufford et al 2013, Ellstrand et al 2007, van Heerwaarden et al 2011}. \jri{check these refs. Hufford 2013 is mexicana, as is i think Ellstrand. do we show introgression with parv in van Heerwaarden?. do is how it in my 2009 paper?}

The lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D in the San Lorenzo population are consistent with patterns expected under neutrality. \jri{lower diversity is consistent with neutrality?} Though, deviations from expected genotype frequencies under Hardy-Weinberg equilibrium at the Hopscotch insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus. \jri{shouldn't Tanja's data be able to argue one way or the other here? she looked at eveidence of inbreeding. what did she find in san lorenzo?} Similarly, if the Hopscotch is segregating as neutral variation in teosinte, the probability that it would have simultaneously drifted to high frequency in multiple populations is quite low; however, we find no evidence of selection or introgression at or near the \emph{tb1} locus in our sample set.

The phenotypic effects of the Hopscotch insertion in domesticated maize have been well documented \cite{Clark et al 2006, Studer et al 2011}, and Weber et al. (2007) have described its effects in partially inbred lines of teosinte \cite{Weber et al 2007}. Our study is the first to explicitly examine the phenotypic effects of the Hopscotch insertion in individuals sampled from a natural population of teosinte. However, we found no significant effect of the Hopscotch on tillering index or tiller number in our phenotyping experiments, and the effect of the Hopscotch insertion in teosinte is discordant with that of maize. The lack of correlation between Hopscotch genotype and tillering index or tiller number is surprising given its effects in maize, as well as our own preliminary data collected on a grow-out of teosinte inbred lines. \jri{Huff, think we should include the TIL data just to show we can see effects in our greenhouse?}

It is certainly possible that even though data from the TILs suggest an effect of the Hopscotch on tillering in teosinte, that it in fact has no effect. \jri{i don't think we want to argue that tb1 has no effect!! we have no evidence for this. instead, it may be that the effect is hard to see if there are numerous other loci and/or epistasis, etc.} Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \cite{Clark et al 2006}, and perhaps we would have detected an effect of the Hopscotch on these additional phenotypes. \jri{why say we might have detected these? we didn't look for them, not sure worth mentioning}  A recent study by Studer and Doebley (2012) examined the possibility of an allelic series at the \emph{tb1} locus in teosinte by introgressing 9 separate teosinte segments spanning the \emph{tb1} locus into an isogenic maize background and investigating their effects on phenotype. However, they found evidence for an allelic series contributing to all \emph{tb1} associated phenotypes but tillering . They did find variation in tillering both between introgressed lines and between isogenic maize lines \jri{don't follow. you mean the different alleles didn't have different tillering phenotypes? but they did show that all alleles affected tillering, right?}, suggesting that there may be additional loci that contribute to tillering in both maize and teosinte \cite{Studer and Doebley 2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering. \jri{are there other studies on other loci? what about Briggs? original Doebley QTL papers? one of the weber papers maybe? seems like there should be explicit cites to other loci we know are involved.  also we could cite \% variance explained by tb1 of tillering in the maize x teo crosses, arguing that there must be other loci.}

Another explanation for the lack of correlation between Hopscotch genotype and tillering could be that growing conditions in our greenhouse affected growth habit of our plants. The \emph{tb1} gene is part of the shade avoidance pathway in maize, and this involvement may have presented problems for our phenotyping design \citep{Kebrom and Brutnell 2007}. Broadly, shade avoidance syndrome (SAS), refers to the morphological and physiological response of plants experiencing an increase in far-red (FR) light, which occurs during shading \cite{Kebrom and Brutnell 2007}. Under increased FR light conditions plants experience both morphological and physiological changes including reduction in branching, increased height, and increased ethylene and auxin production \cite{Ballare 1999, Smith and Whitelam 1997}. In maize, increase in shading has been shown to be correlated with changes in \emph{tb1} expression and subsequent reduction in tillering, suggesting that \emph{tb1} is involved in the shade avoidance response in maize \cite{Doebley et al 1997, Lukens and Doebley 1999}. If plants in our phenotyping experiments were grown too close together, such that they shaded each other, it could have reduced \emph{tb1} expression and tillering, which would have confounded our results. \jri{maybe, but we saw variation in tillering, especially in growout number 1. shade would have to so effect expression that it didn't matter what tb1 genotype a plant had -- so G x E effects.  is it known whether there are such G x E effects?}

In summary, our findings demonstrate that the Hopscotch allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the Hopscotch allele is present in these populations due to recent introgression; however, it seems unlikely that the it would have drifted to high frequency in multiple populations, and there may be another explanation for the high frequency we observe in some of our populations. The Hopscotch insertion does not appear to have any affect on tillering phenotypes in teosinte previously associated with \emph{tb1} and it is possible that variation at other loci may play a larger role in branching architecture in teosinte than in domesticated maize. \jri{seems like we should end with at least a hypothesis. as is we are saying we ruled out drift, introgression, and selection. what else is there? perhaps we should think about selection or introgression scenarios that could explain our data.}

\clearpage
\bibliography{refs}
